Constraining gravitational wave amplitude birefringence with GWTC-3

### 2023-10-17 src/tex/ms.tex
Update changes from PRD

\usepackage{showyourwork}
\usepackage{amsfonts,amssymb,amsmath}
\usepackage[nolist,nohyperlinks]{acronym}
\usepackage{bookmark}
\usepackage{changes}
%\usepackage[all]{hypcap}

@@

\newcommand*{\mi}[1]{\textsf{\color{magenta} [\textbf{MAX:} #1]}}
\newcommand*{\wf}[1]{\textsf{\color{cyan} [\textbf{WILL:} #1]}}

\definechangesauthor[name=Will, color=cyan]{WF}

@@

\author{Maximiliano Isi}
\email{misi@flatironinstitute.org}
\affiliation{Center for Computational Astrophysics, Flatiron Institute, 162 5th Ave, New York, NY 10010, United States}

\author{Kaze W. K. Wong}
\email{kwong@flatironinstitute.org}
\affiliation{Center for Computational Astrophysics, Flatiron Institute, 162 5th Ave, New York, NY 10010, United States}

\author{Will M. Farr}
\email{wfarr@flatironinstitute.org}
\affiliation{Center for Computational Astrophysics, Flatiron Institute, 162 5th Ave, New York, NY 10010, United States}
\affiliation{Department of Physics and Astronomy, Stony Brook University, Stony Brook NY 11794, United States}

@@

\begin{abstract}
    The propagation of gravitational waves can reveal fundamental features of the structure of spacetime.
    For instance, differences in the propagation of gravitational-wave polarizations would be a smoking gun for parity violations in the gravitational sector, as expected from birefringent theories like Chern-Simons gravity.
    Here we look for evidence of amplitude birefringence in the latest LIGO-Virgo catalog (GWTC-3) through the use of birefringent templates inspired by dynamical Chern-Simons gravity.
    From 71 binary-black-hole signals, we obtain the most precise constraints on gravitational-wave amplitude birefringence yet, measuring a birefringent attenuation of \variable{output/restricted_kappa_median.txt} at $100 \, \mathrm{Hz}$ with 90\% credibility, equivalent to a parity-violation energy scale of \variable{output/M_PV_constraint.txt}.
\end{abstract}

@@

\begin{acronym}
\acro{GW}{gravitational wave}
\acro{GR}{general relativity}
\acro{CBC}{compact-binary coalescence}
\acro{BH}{black hole}
\acro{BBH}{binary black hole}
\acro{LVK}{LIGO-Virgo-KAGRA Collaboration}
\acro{PE}{parameter estimation}
\acro{FAR}{false-alarm rate}
\acro{GWOSC}{the Gravitational Wave Open Science Center}
\acro{SNR}{signal-to-noise ratio}
\end{acronym}

@@

% Previous studies
Previous studies have constrained amplitude birefringence by performing different statistical analyses.
\citet{Yamada_2020} and \citet{Wang_2021} both performed \ac{PE} on the events in the first \ac{GW} transient catalog \citep{GWTC-1}, GWTC-1, using birefringent templates.
\citet{Okounkova_2022} considered the distribution of observed inclinations of the \ac{GW} events in the second \ac{GW} transient catalog \citep{GWTC-2}, GWTC-2, to look for signs of birefringence; \citet{Vitale:2022pmu} carried out a related analysis.
While this manuscript was finalized, \citet{Zhu:2023wci} reported an analysis of GWTC-3.

% What's new?
In this study, we use a frequency-dependent birefringence model to constrain the strength of \ac{GW} amplitude birefringence by performing \ac{PE} on \ac{LVK} binaries.
This model is a better approximation of the birefringence effect expected from theory than the frequency-independent model used in \citet{Okounkova_2022}.
\added{Including the frequency dependence allows us to break the degeneracy between birefringence and source inclination, which we discuss below in Sec.~\ref{sec:inclination}.}
Compared to \replaced{Refs.~\citep{Yamada_2020,Wang_2021,Okounkova_2022}}{other studies}, we perform \ac{PE} on more events, including events new to GWTC-3 \citep{GWTC-3}, and use a phenomenology-oriented parametrization.
We consider 71 binary black hole merger events with a \ac{FAR} $\leq1/\mathrm{yr}$, as listed in Table I of \citet{GWTC-3_population}.
We discuss single-event results in detail, and identify degeneracies between birefringence and spin effects, in addition to the already known correlations with source orientation and distance.
We use the results from individual events to place a collective population constraint on the strength of \ac{GW} amplitude birefringence from GWTC-3.

% Section guide
In Sec.~\ref{sec:Background}, we briefly review the background of \ac{GW} amplitude birefringence.
In Sec.~\ref{sec:Method}, we describe the modification we apply to the baseline \ac{GR} waveform, summarize our \ac{PE} configuration and outline our statistical methods for combining information from multiple events.
In Sec.~\ref{sec:Results}, we present our constraint on \ac{GW} amplitude birefringence, discussing individual events and the catalog collectively.
In Sec.~\ref{sec:Discussion}, we discuss the implications of our result, comparing to previous constraints in the literature and outlining correlation structures that appear in our measurements.
Finally, we conclude in Sec.~\ref{sec:Discussion} with a summary.

@@

% GW polarization
In \ac{GR}, \acp{GW} are comprised of two independent polarization modes, usually represented in the linear basis of plus ($+$) and cross ($\times$) states.
In the Fourier domain, these can be combined into left-handed (L) and right-handed (R) circular states (see, e.g., \cite{Isi:2022mbx}),
\begin{equation}
    h_{L/R} = \frac{1}{\sqrt{2}}\left(h_+ \pm i h_\times\right)\,,
\end{equation}
where $h$ is the frequency domain strain, with the plus and minus signs for L and R respectively.
These circular modes represent eigenstates of the helicity operator (helicity $\pm2$) and possess a definite parity.
Einstein's theory, which is parity even, predicts no difference in the dynamics of these two states.

@@

In particular, \emph{amplitude} birefringence would enhance one polarization mode over the other.
\added{Following \cite{Alexander:2009tp}, the effect of birefringence can be modeled as a frequency-dependent amplification or dampening; similar derivations can be found in \cite{Zhao:2019xmm, Ezquiaga:2021ler, Zhu:2023wci}.}
To first order, in theories like Chern-Simons gravity, the Fourier-domain waveform observed a comoving distance $d_C$ away from the source can be written as
\begin{equation}
    h_{L/R}^{\mathrm{br}}(f) =
    h_{L/R}^{\mathrm{GR}}(f) \times
    \exp\left(\pm\kappa d_C \frac{f}{100\,\mathrm{Hz}}\right)\,,
    \label{eq:waveform_modification}
\end{equation}
where the emitted waveform $h_{L/R}^{\mathrm{GR}}$ is modified by an exponential birefringent factor to yield the observed waveform $h_{L/R}^{\mathrm{br}}$.
The overall magnitude of this effect for a given frequency $f$ is set by an attenuation coefficient, $\kappa$, which encodes the intrinsic strength of the birefringence.
The emitted waveform for a given source (i.e., the waveform observed in the near zone, very close to the source) will generally differ from the analogous waveform predicted by \ac{GR} \cite{Alexander:2009tp,Okounkova:2019zjf}; however, since we expect most viable modifications to \ac{GR} to be intrinsically small (e.g., \cite{Okounkova:2022grv}), it is standard to approximate the emitted waveform by the prediction from \ac{GR} (hence the notation ``$h^{\rm GR}$'' above).

@@

Equation \eqref{eq:waveform_modification} can be derived as the first order effect in an expansion away from \ac{GR} under multiple frameworks.
In general, $\kappa$ will be a function of the theory parameters and the cosmological history, e.g., the value of the pseudo-scalar field and its derivative in Chern-Simons gravity \cite{Alexander:2009tp}.
Since it originates from a truncated series expansion%
\footnote{Often the birefringent effect is written as an expansion in redshift, $z$ \cite[e.g.][]{Zhao:2019xmm}, rather than distance.  Here we use $d_C$ to emphasize that the effect accumulates \emph{per cycle}, or with propagation distance.  Thus, current ground-based experiments, which can detect gravitational waves to $z \simeq 1$, $d_C \simeq d_H$ (Hubble distance), are already observing propagation over a large fraction of the universe and constrain this effect meaningfully.}, %
Eq.~\eqref{eq:waveform_modification} is a good approximation only for small exponents, 
\begin{equation}
    \left|\kappa\right| \left(d_c\added{/1\, \mathrm{Gpc}}\right) \left(f/100\, \mathrm{Hz}\right) < 1\, \added{\mathrm{Gpc}^{-1}}\,.
    \label{eq:small_exponent}
\end{equation}
Otherwise, more frequency-dependent terms would enter the exponent of Eq.~\eqref{eq:waveform_modification} in a theory-dependent way.
\begin{figure}
    \script{birefringence.py}
    \includegraphics[width=\columnwidth]{figures/birefringence.pdf}
    \caption{
        \emph{Illustration of amplitude birefringence.} The GR waveform for the $\ell=|m|=2$ mode of a nonprecessing BBH seen edge-on ($\cos\iota = 0$) is linearly polarized and thus contains equal amounts of left- and right-handed modes for all frequencies (dotted, top).
        However, if spacetime were birefringent following Eq.~\protect\eqref{eq:waveform_modification}, the waveform observed on Earth would contain different fractions of the two circular modes, with higher frequencies affected more strongly (solid, top). 
        In the time domain, this manifests as a time-dependent amplification of the waveform, with a stronger effect at later times when the chirp reaches a higher instantaneous frequency (bottom).
        For this example, the black holes do not spin and have equal masses $m_1 = m_2 = 10\, M_\odot$, and we have chosen a luminosity distance $d_L = 400\, {\rm Mpc}$ and $\kappa = 0.6$.
        }
    \label{fig:birefringence}
\end{figure}

@@

% Effect on inclination PE results
Under certain conditions, the effect of birefringence can be degenerate with a change in the orientation of the source with respect to the line of sight \cite{Alexander:2009tp}.
Concretely, for a nonprecessing compact binary inspiral in \ac{GR}, the observed amplitude ratio of the left-handed and right-handed polarizations is only a function of the inclination $\iota$, the angle between the orbital angular momentum of the source and the line of sight \cite{Blanchet:2013haa}.
For  the dominant $\ell = |m| = 2$ angular mode of the radiation, the relation between the amplitude ratio and the inclination is
\begin{equation}
    \frac{h_{L}^\mathrm{GR}}{h^\mathrm{GR}_{R}}=\left(\frac{1-\cos\iota}{1+\cos\iota}\right)^2\,
\end{equation}
for all frequencies (see, e.g., Sec.~IIIC in \cite{Isi:2022mbx}).

Since birefringence impacts the observed amplitude ratio of left- and right-handed modes, it could also affect inferences about the source inclination \cite{Alexander:2009tp}.
However, the two effects are degenerate only if the frequency dependence of Eq.~\eqref{eq:waveform_modification} is neglected.
This is easy to see from Eq.~\eqref{eq:waveform_modification}, since the implied polarization ratio for the $\ell = |m| = 2$ mode of a nonprecessing source is
\begin{equation}
    \frac{h_{L}^\mathrm{br}}{h_{R}^\mathrm{br}}=\left(\frac{1-\cos\iota}{1+\cos\iota}\right)^2
    \exp\left(2\kappa d_C \frac{f}{100\, \mathrm{Hz}}\right)\, .
    \label{eq:modified_amplitude_ratio}
\end{equation}
For an isolated Fourier mode of definite frequency $f$, the effect of birefringence will be degenerate with a change in inclination; however, if multiple modes come into play, then no change in inclination alone can mask the effect of birefringence, which will affect the time domain waveform nontrivially (Fig.~\ref{fig:birefringence}).

@@

By implementing Eq.~\eqref{eq:waveform_modification}, we generally break the degeneracy between birefringence and source orientation; this was also the case in the frequency-dependent relations studied in \cite{Yamada_2020,Wang_2021}.
Nevertheless, there exist systems for which the degeneracy cannot be broken in practice because not enough frequencies are available in the data.
This may also be the case for quasimonochromatic sources, like nonaxisymmetric pulsars or very light binaries, which are well approximated by a single Fourier mode.

@@

To constrain birefringence, we reanalyze events from GWTC-3 \citep{GWTC-2.1, GWTC-3} implementing Eq.~\eqref{eq:waveform_modification} to directly obtain a posterior on $\kappa$ from the strain of each event.
We analyze the 71 \acp{BBH} that were detected with $\mathrm{FAR} < 1/\mathrm{yr}$\added{, less stringent than the typical \ac{LVK} threshold of $1/1000\,\mathrm{yr}$ \cite{LIGOScientific:2020tif,LIGOScientific:2021sio}.
The \ac{FAR} values are typically determined by \ac{GR} pipelines, which could down-rank signals beyond \ac{GR} \cite{LIGOScientific:2020tif,Chia:2020psj,tgrsel}; however, since detectability through matched-filtering is most sensitive to the phase, not the amplitude, we expect only a minor decrease in sensitivity to the kind of birefringent signals explored here.}
To avoid extended computations on longer signals and considering these are generally at closer distances, we do not analyze systems involving neutron stars in this work.
We procure strain data from \ac{GWOSC} \citep{GWOSC}.

We estimate source parameters using a custom version of the \textsc{Bilby} software \citep{Bilby}, modified from the baseline version to apply Eq.~\eqref{eq:waveform_modification} for any \ac{GR} baseline waveform.
We take the \ac{PE} configuration in \citep{GWTC-2.1, GWTC-3, GWTC-2.1_dataset, GWTC-3_dataset} as a starting point, with \textsc{IMRPhenomXPHM} \citep{Pratten:2020ceb} as the reference waveform.
We apply a distance prior corresponding to a uniform distribution over comoving volume and source-frame time, and set the prior on $\kappa$ to be uniform between $-1 \, \mathrm{Gpc}^{-1}$ and $1 \, \mathrm{Gpc}^{-1}$.

@@

In the most simplified scenarios, birefringence is a property of spacetime that is not intrinsic to any source or region in space. 
Consequently, we should expect $\kappa$ to take the same value for all signals, whether it vanishes or not.
Under this assumption, the posterior on $\kappa$ inferred collectively from all events is simply obtained from the product of the individual likelihoods, $p(d_i \mid\kappa) \propto p(\kappa \mid d_i)/p(\kappa)$, such that
\begin{equation}
    p(\kappa \mid \{d_i\})\propto p(\kappa) \prod_{i}\frac{p(\kappa \mid d_i)}{p(\kappa)}\,,
    \label{eq:restricted_posterior}
\end{equation}
where $d_i$ is the strain data for the $i$\textsuperscript{th} event, and $p(\kappa)$ is the prior on $\kappa$; since the prior is uniform, in our case Eq.~\eqref{eq:restricted_posterior} reduces to the product of the posteriors, namely $p(\kappa \mid \{d_i\}) \propto \prod_{i}p(\kappa \mid d_i)$.
We use Eq.~\eqref{eq:restricted_posterior} to obtain the primary constraint presented in this work.

@@

To do this, we apply hierarchical Bayesian inference \cite{Mandel2010,Hogg2010} to model the distribution of $\kappa$'s consistent with the observed data:
we posit that, rather than a unique global value of $\kappa$, there is a specific value of the parameter, $\kappa_i$, associated with each event, and that this is drawn from some unknown distribution of true underlying values; from the imperfect measurements of $\kappa_i$ for each event, we may reconstruct the underlying distribution.
If we are interested in constraining the first two moments of the distribution, it is convenient to parametrize the $\kappa_i$'s as drawn from a Gaussian with unknown mean $\mu$ and variance $\sigma^2$, i.e., $\kappa_i \sim \mathcal{N}(\mu, \sigma^2)$ \cite{Isi:2019asy}, and measure those hyperparameters from the collection of observed data.

If \ac{GR} is correct and there is no birefringence, we should find the observed $\kappa$ distribution to be consistent with a delta function at the origin ($\kappa_i = 0$ for all $i$, or $\mu=\sigma=0$); on the other hand, if spacetime is globally birefringent, we expect to find a delta function at some nonzero value ($\kappa_i = \kappa \neq 0$, or $\mu = \kappa$ and $\sigma=0$).
But this analysis also has the power to reveal unexpected physics or systematics in our measurements: if $\sigma$ is confidently found to be nonzero, this would imply that our set of measurements is statistically unlikely to originate from a unique $\kappa$ value.
This could signal richer physics than is implied by Eq.~\eqref{eq:waveform_modification} or, more prosaically, that there are outliers in our measurements due to mismodeling, e.g., in the waveform approximant or the noise of the detector.

@@

In this section, we present the results of our study.
We first show the $\kappa$ measurements of all events in our set, as well as the resulting global measurement of $\kappa$ that represents our primary constraint on birefringence (Sec.~\ref{sec:results:gwtc}).
We then assess the collection of measurements in more detail through a hierarchical analysis (Sec.~\ref{sec:results:hier}).
Finally, we discuss some special events individually, and outline the degeneracies that arise between birefringence and orbital precession (Sec.~\ref{sec:results:notable}).

\begin{figure}
    \script{violin_kappa.py}
    \includegraphics[width=\columnwidth]{figures/violin_kappa.pdf}
    \caption{
        Individual-event $\kappa$ posteriors (distributions), and joint measurement (blue band, 90\% CI; blue line, median).
    }
    \label{fig:violin_kappa}
\end{figure}

@@

Figure \ref{fig:violin_kappa} makes it clear that not all \acp{BBH} in GWTC-3 are equally informative about birefringence.
When considered individually, the events that best constrain $\kappa$ are listed in Table~\ref{tab:best_events_kappa}, in order of increasing standard deviation $\sigma_i$.
That table also shows the credible level (CL) at which the posterior supports $\kappa = 0$, whereby $\mathrm{CL} = 0$ ($\mathrm{CL} = 1$) means the posterior supports that value with high (low) probability.
\added{This quantity can be understood as the relative height of the probability density function at a given value, and does not assume any symmetric interval or modality.}

@@

To this end, we perform three additional \ac{PE} runs for GW200129, considering only two detectors at a time: LIGO Hanford and Virgo (HL), LIGO Livingston and Virgo (LV), and LIGO Hanford and LIGO Livingston (HL).
If the preference for $\kappa < 0$ is tied to the glitch in Virgo, we expect it to disappear in the HL run, which excludes Virgo data.

>>>>

\usepackage{showyourwork}
\usepackage{amsfonts,amssymb,amsmath}
\usepackage[nolist,nohyperlinks]{acronym}
\usepackage{bookmark}
\usepackage{changes}

@@

@@

\author{Maximiliano Isi}
\email{misi@flatironinstitute.org}
\affiliation{Center for Computational Astrophysics, Flatiron Institute, 162 5th Avenue, New York, New York 10010, USA}

\author{Kaze W. K. Wong}
\email{kwong@flatironinstitute.org}
\affiliation{Center for Computational Astrophysics, Flatiron Institute, 162 5th Avenue, New York, New York 10010, USA}

\author{Will M. Farr}
\email{wfarr@flatironinstitute.org}
\affiliation{Center for Computational Astrophysics, Flatiron Institute, 162 5th Avenue, New York, New York 10010, USA}
\affiliation{Department of Physics and Astronomy, Stony Brook University, Stony Brook, New York 11794, USA}

@@

\begin{abstract}
    The propagation of gravitational waves can reveal fundamental features of the structure of spacetime.
    For instance, differences in the propagation of gravitational-wave polarizations would be a smoking gun for parity violations in the gravitational sector, as expected from birefringent theories like Chern-Simons gravity.
    Here we look for evidence of amplitude birefringence in the \replaced{third gravitational-wave transient}{latest Laser Interferometer Gravitational Wave Observatory-Virgo} catalog (GWTC-3) through the use of birefringent templates inspired by dynamical Chern-Simons gravity.
    From 71 binary-black-hole signals, we obtain the most precise constraints on gravitational-wave amplitude birefringence yet, measuring a birefringent attenuation of \variable{output/restricted_kappa_median.txt} at $100 \, \mathrm{Hz}$ with 90\% credibility, equivalent to a parity-violation energy scale of \variable{output/M_PV_constraint.txt}.
\end{abstract}

@@

\begin{acronym}
\acro{GW}{gravitational wave}
\acro{GR}{general relativity}
\acro{CBC}{compact-binary coalescence}
\acro{BH}{black hole}
\acro{BBH}{binary black hole}
\acro{LVK}{LIGO-Virgo-KAGRA Collaboration}
\acro{PE}{parameter estimation}
\acro{FAR}{false-alarm rate}
\acro{SNR}{signal-to-noise ratio}
\acro{2D}{two-dimensional}
\end{acronym}

@@

% Previous studies
Previous studies have constrained amplitude birefringence by performing different statistical analyses.
\citet{Yamada_2020}, \citet{Wang_2021} both performed \ac{PE} on the events in the first \ac{GW} transient catalog \citep{GWTC-1}, GWTC-1, using birefringent templates.
\citet{Okounkova_2022} considered the distribution of observed inclinations of the \ac{GW} events in the second \ac{GW} transient catalog \citep{GWTC-2}, GWTC-2, to look for signs of birefringence; \citet{Vitale:2022pmu} carried out a related analysis.
While this manuscript was finalized, \citet{Zhu:2023wci} reported an analysis of GWTC-3.

% What's new?
In this study, we use a frequency-dependent birefringence model to constrain the strength of \ac{GW} amplitude birefringence by performing \ac{PE} on \ac{LVK} binaries.
This model is a better approximation of the birefringence effect expected from theory rather than the frequency-independent model used in \citet{Okounkova_2022}.
Including the frequency dependence allows us to break the degeneracy between birefringence and source inclination, which we discuss below in Sec.~\ref{sec:inclination}.
Compared to Refs.~\citep{Yamada_2020,Wang_2021,Okounkova_2022}, we perform \ac{PE} on more events, including events new to GWTC-3 \citep{GWTC-3}, and use a phenomenology-oriented parametrization.
We consider 71 binary black hole merger events with a \ac{FAR} $\leq1/\mathrm{yr}$, as listed in Table I of \citet{GWTC-3_population}.
We discuss single-event results in detail, and identify degeneracies between birefringence and spin effects, in addition to the already known correlations with source orientation and distance.
We use the results from individual events to place a collective population constraint on the strength of \ac{GW} amplitude birefringence from GWTC-3.

% Section guide
In Sec.~\ref{sec:Background}, we briefly review the background of \ac{GW} amplitude birefringence.
In Sec.~\ref{sec:Method}, we describe the modification we apply to the baseline \ac{GR} waveform, summarize our \ac{PE} configuration and outline our statistical methods for combining information from multiple events.
In Sec.~\ref{sec:Results}, we present our constraint on \ac{GW} amplitude birefringence, discussing individual events and the catalog collectively.
In Sec.~\ref{sec:Discussion}, we discuss the implications of our result, comparing to previous constraints in the literature and outlining correlation structures that appear in our measurements.
Finally, we conclude in Sec.~\replaced{\ref{sec:Conclusion}}{\ref{sec:Discussion}} with a summary.

@@

% GW polarization
In \ac{GR}, \acp{GW} are comprised of two independent polarization modes, usually represented in the linear basis of plus ($+$) and cross ($\times$) states.
In the Fourier domain, these can be combined into left-handed (\textit{L}) and right-handed (\textit{R}) circular states (see, e.g., \cite{Isi:2022mbx}),
\begin{equation}
    h_{L/R} = \frac{1}{\sqrt{2}}\left(h_+ \pm i h_\times\right)\,,
\end{equation}
where $h$ is the frequency domain strain, with the plus and minus signs for L and R respectively.
These circular modes represent eigenstates of the helicity operator (helicity $\pm2$) and possess a definite parity.
Einstein's theory, which is parity even, predicts no difference in the dynamics of these two states.

@@

In particular, \emph{amplitude} birefringence would enhance one polarization mode over the other.
Following \cite{Alexander:2009tp}, the effect of birefringence can be modeled as a frequency-dependent amplification or dampening; similar derivations can be found in \cite{Zhao:2019xmm, Ezquiaga:2021ler, Zhu:2023wci,Jenks:2023pmk}.
To first order in the equations of motion, in theories like Chern-Simons gravity, the Fourier-domain waveform observed at a comoving distance $d_C$ away from the source can be written as
\begin{equation}
    h_{L/R}^{\mathrm{br}}(f) =
    h_{L/R}^{\mathrm{GR}}(f) \times
    \exp\left(\pm\kappa\, d_C \frac{f}{100\,\mathrm{Hz}}\right)\,,
    \label{eq:waveform_modification}
\end{equation}
where the emitted waveform $h_{L/R}^{\mathrm{GR}}$ is modified by an exponential birefringent factor to yield the observed waveform $h_{L/R}^{\mathrm{br}}$.
The overall magnitude of this effect for a given frequency $f$ is set by an attenuation coefficient $\kappa$, which encodes the intrinsic strength of the birefringence:
$\kappa^{-1}$ represents an ``attenuation length'' encoding the typical distance that yields an $e$-folding in the amplification or dampening of the polarizations at a fiducial signal frequency of 100 Hz at the detector.
The emitted waveform for a given source (i.e., the waveform observed in the near zone, very close to the source) will generally differ from the analogous waveform predicted by \ac{GR} \cite{Alexander:2009tp,Okounkova:2019zjf}; however, since we expect most viable modifications to \ac{GR} to be intrinsically small (e.g., \cite{Okounkova:2022grv}), it is standard to approximate the emitted waveform by the prediction from \ac{GR} (hence the notation ``$h^{\rm GR}$'' above).

@@

Equation \eqref{eq:waveform_modification} can be derived as the first-order effect in an expansion away from \ac{GR} under multiple frameworks.
In general, $\kappa$ will be a function of the theory parameters and the cosmological history, e.g., the value of the pseudoscalar field and its derivative in Chern-Simons gravity \cite{Alexander:2009tp}.
Since it originates from a truncated series expansion,%
\footnote{Often the birefringent effect is written as an expansion in redshift $z$ \cite[e.g.][]{Zhao:2019xmm}, rather than distance. Here we use $d_C$ to emphasize that the effect accumulates \emph{per cycle}, or with propagation distance.  Thus, current ground-based experiments, which can detect gravitational waves to $z \simeq 1$, $d_C \simeq d_H$ (Hubble distance), are already observing propagation over a large fraction of the Universe and constrain this effect meaningfully.} %
Eq.~\eqref{eq:waveform_modification} is a good approximation only for small exponents, namely
\begin{equation}
    \left(\frac{\left|\kappa\right|}{\mathrm{Gpc}^{-1}}\right) \left(\frac{d_c}{\mathrm{Gpc}}\right) \left(\frac{f}{100\, \mathrm{Hz}}\right) < 1\,,
    \label{eq:small_exponent}
\end{equation}
recalling that $\kappa$ has dimensions of inverse length.
Otherwise, more frequency-dependent terms could enter the exponent of Eq.~\eqref{eq:waveform_modification} in a theory-dependent way.
\begin{figure}
    \script{birefringence.py}
    \includegraphics[width=\columnwidth]{figures/birefringence.pdf}
    \caption{
        Illustration of amplitude birefringence. The GR waveform for the $\ell=|m|=2$ mode of a nonprecessing BBH seen edge-on ($\cos\iota = 0$) is linearly polarized and thus contains equal amounts of left- and right-handed modes for all frequencies (dotted, top).
        However, if spacetime were birefringent following Eq.~\protect\eqref{eq:waveform_modification}, then the waveform observed on Earth would contain different fractions of the two circular modes, with higher frequencies affected more strongly (solid, top). 
        In the time domain, this manifests as a time-dependent amplification of the waveform, with a stronger effect at later times when the chirp reaches a higher instantaneous frequency (bottom).
        For this example, the black holes do not spin and have equal masses $m_1 = m_2 = 10\, M_\odot$, and we have chosen a luminosity distance $d_L = 400\, {\rm Mpc}$ and $\kappa = 0.6$.
        }
    \label{fig:birefringence}
\end{figure}

@@

% Effect on inclination PE results
Under certain conditions, the effect of birefringence can be degenerate with a change in the orientation of the source with respect to the line of sight \cite{Alexander:2009tp}.
Concretely, for a nonprecessing compact binary inspiral in \ac{GR}, the observed amplitude ratio of the left-handed and right-handed polarizations is only a function of the inclination $\iota$, the angle between the orbital angular momentum of the source and the line of sight \cite{Blanchet:2013haa}.
For  the dominant $\ell = |m| = 2$ angular mode of the radiation, the relation between the amplitude ratio and the inclination is
\begin{equation}
    \frac{h_{L}^\mathrm{GR}}{h^\mathrm{GR}_{R}}=\left(\frac{1-\cos\iota}{1+\cos\iota}\right)^2\,
\end{equation}
for all frequencies (see, e.g., Sec.~III C in \cite{Isi:2022mbx}).

Since birefringence impacts the observed amplitude ratio of left- and right-handed modes, it could also affect inferences about the source inclination \cite{Alexander:2009tp}.
However, the two effects are degenerate only if the frequency dependence of Eq.~\eqref{eq:waveform_modification} is neglected.
This is easy to see from Eq.~\eqref{eq:waveform_modification} since the implied polarization ratio for the $\ell = |m| = 2$ mode of a nonprecessing source is
\begin{equation}
    \frac{h_{L}^\mathrm{br}}{h_{R}^\mathrm{br}}=\left(\frac{1-\cos\iota}{1+\cos\iota}\right)^2
    \exp\left(2\kappa d_C \frac{f}{100\, \mathrm{Hz}}\right)\, .
    \label{eq:modified_amplitude_ratio}
\end{equation}
For an isolated Fourier mode of definite frequency $f$, the effect of birefringence will be degenerate with a change in inclination; however, if multiple modes come into play, then no change in inclination alone can mask the effect of birefringence, which will affect the time domain waveform nontrivially (Fig.~\ref{fig:birefringence}).

@@

By implementing Eq.~\eqref{eq:waveform_modification}, we generally break the degeneracy between birefringence and source orientation; this was also the case in the frequency-dependent relations studied in \cite{Yamada_2020,Wang_2021}.
Nevertheless, there exist systems for which the degeneracy cannot be broken, in practice, because not enough frequencies are available in the data.
This may also be the case for quasimonochromatic sources, like nonaxisymmetric pulsars or very light binaries, which are well approximated by a single Fourier mode.

@@

To constrain birefringence, we reanalyze events from GWTC-3 \citep{GWTC-2.1, GWTC-3} implementing Eq.~\eqref{eq:waveform_modification} to directly obtain a posterior on $\kappa$ from the strain of each event.
We analyze the 71 \acp{BBH} that were detected with $\mathrm{FAR} < 1/\mathrm{yr}$, less stringent than the typical \ac{LVK} threshold of $1/1000\,\mathrm{yr}$ \cite{LIGOScientific:2020tif,LIGOScientific:2021sio}.
The \ac{FAR} values are typically determined by \ac{GR} pipelines, which could down-rank signals beyond \ac{GR} \cite{LIGOScientific:2020tif,Chia:2020psj,tgrsel}; however, since detectability through matched-filtering is most sensitive to the phase, not the amplitude, we expect only a minor decrease in sensitivity to the kind of birefringent signals explored here.
To avoid extended computations on longer signals and considering these are generally at closer distances, we do not analyze systems involving neutron stars in this work.
We procure strain data from \add{the} Gravitational Wave Open Science Center \citep{GWOSC}.

We estimate source parameters using a custom version of the \textsc{Bilby} software \citep{Bilby}, modified from the baseline version to apply Eq.~\eqref{eq:waveform_modification} for any \ac{GR} baseline waveform.
We take the \ac{PE} configuration in \citep{GWTC-2.1, GWTC-3, GWTC-2.1_dataset, GWTC-3_dataset} as a starting point, with \textsc{IMRPhenomXPHM} \citep{Pratten:2020ceb} as the reference waveform.
We apply a distance prior corresponding to a uniform distribution over comoving volume and source-frame time (see, e.g., Eq.~10 in \cite{LIGOScientific:2019zcs}), and set the prior on $\kappa$ to be uniform between $-1 \, \mathrm{Gpc}^{-1}$ and $1 \, \mathrm{Gpc}^{-1}$.

@@

In the most simplified scenarios, birefringence is a property of spacetime that is not intrinsic to any source or region in space. 
Consequently, we should expect $\kappa$ to take the same value for all signals, whether it vanishes or not.
Under this assumption, the posterior on $\kappa$ inferred collectively from all events is simply obtained from the product of the individual likelihoods, $p(d_i \mid\kappa) \propto p(\kappa \mid d_i)/p(\kappa)$, such that
\begin{equation}
    p(\kappa \mid \{d_i\})\propto p(\kappa) \prod_{i}\frac{p(\kappa \mid d_i)}{p(\kappa)}\,,
    \label{eq:restricted_posterior}
\end{equation}
where $d_i$ is the strain data for the $i$th event, and $p(\kappa)$ is the prior on $\kappa$; since the prior is uniform, in our case Eq.~\eqref{eq:restricted_posterior} reduces to the product of the posteriors, namely $p(\kappa \mid \{d_i\}) \propto \prod_{i}p(\kappa \mid d_i)$.
We use Eq.~\eqref{eq:restricted_posterior} to obtain the primary constraint presented in this work.

@@

To do this, we apply hierarchical Bayesian inference \cite{Mandel2010,Hogg2010} to model the distribution of $\kappa$'s consistent with the observed data:
we posit that, rather than a unique global value of $\kappa$, there is a specific value of the parameter, $\kappa_i$, associated with each event, and that this is drawn from some unknown distribution of true underlying values; from the imperfect measurements of $\kappa_i$ for each event, we may reconstruct the underlying distribution.
If we are interested in constraining the first two moments of the distribution, then it is convenient to parametrize the $\kappa_i$'s as drawn from a Gaussian with unknown mean $\mu$ and variance $\sigma^2$, i.e., $\kappa_i \sim \mathcal{N}(\mu, \sigma^2)$ \cite{Isi:2019asy}, and measure those hyperparameters from the collection of observed data.

If \ac{GR} is correct and there is no birefringence, then we should find the observed $\kappa$ distribution to be consistent with a delta function at the origin ($\kappa_i = 0$ for all $i$, or $\mu=\sigma=0$); on the other hand, if spacetime is globally birefringent, then we expect to find a delta function at some nonzero value ($\kappa_i = \kappa \neq 0$, or $\mu = \kappa$ and $\sigma=0$).
But this analysis also has the power to reveal unexpected physics or systematics in our measurements: if $\sigma$ is confidently found to be nonzero, then this would imply that our set of measurements is statistically unlikely to originate from a unique $\kappa$ value.
This could signal richer physics than is implied by Eq.~\eqref{eq:waveform_modification} or, more prosaically, that there are outliers in our measurements due to mismodeling, e.g., in the waveform approximant or the noise of the detector.

@@

In this section, we present the results of our study.
We first show the $\kappa$ measurements from all events in our set, as well as the resulting global measurement of $\kappa$ that represents our primary constraint on birefringence (Sec.~\ref{sec:results:gwtc}).
We then assess the collection of measurements in more detail through a hierarchical analysis (Sec.~\ref{sec:results:hier}).
Finally, we discuss some special events, individually, and outline the degeneracies that arise between birefringence and orbital precession (Sec.~\ref{sec:results:notable}).

\begin{figure}
    \script{violin_kappa.py}
    \includegraphics[width=\columnwidth]{figures/violin_kappa.pdf}
    \caption{
        Individual-event $\kappa$ posteriors (distributions) and joint measurement (blue band, 90\% CI; blue line, median).
    }
    \label{fig:violin_kappa}
\end{figure}

@@

Figure \ref{fig:violin_kappa} makes it clear that not all \acp{BBH} in GWTC-3 are equally informative about birefringence.
When considered individually, the events that best constrain $\kappa$ are listed in Table~\ref{tab:best_events_kappa}, in order of increasing standard deviation $\sigma_i$.
That table also shows the credible level (CL) at which the posterior supports $\kappa = 0$, whereby $\mathrm{CL} = 0$ ($\mathrm{CL} = 1$) means the posterior supports that value with high (low) probability.
This quantity represents the relative height of the probability density function at $\kappa = 0$, and is not tied to a symmetric interval, making it particularly useful in assessing bimodal posteriors.

@@

To this end, we perform three additional \ac{PE} runs for GW200129, considering only two detectors at a time: LIGO Hanford and Virgo (\replaced{HV}{HVL}), LIGO Livingston and Virgo (LV), and LIGO Hanford and LIGO Livingston (HL).
If the preference for $\kappa < 0$ is tied to the glitch in Virgo, we expect it to disappear in the HL run, which excludes Virgo data.
